Think deep for reef resilience

### 2025-07-05 paper/paper_draft.qmd
clear whitespace
Revisions for clarity/flow

Climate change is impacting species survival, persistence and ecosystem functions globally, including coral. Coral reefs are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events linked to climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral reefs are, and will continue to, decrease the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well as ecosystem services such as food, protection, tourism and Indigenous cultural value [@eddy2021; @stoeckl2021]. We aim to assess the impacts of possible climate futures and identify the characteristics of resilient reefs using the latest generation of larval connectivity research with scenario-based population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. Severe declines in coral cover are projected due to compounding impacts such as ocean warming and acidification, intensification of Tropical Cyclones, and Crown of Thorns Starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the 2023-24 Summer mass bleaching event ([@cantin2024]) may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs.

Environmental abiotic factors influence how reefs will experience climate change pressures, including the spatio-temporal patterns of thermal stress itself, and indirect factors that influence the stress experienced by corals. Environmental factors such as tropical cyclones and wind driven mixing [@lin2023; @mcclanahan2005]; currents [@grimaldi2022] and depth [@álvarez-noriega2018; @baird2018; @smith2014] can decrease the magnitude of thermal stress corals experience. On the other hand, high light intensity and nutrient levels in the benthic zone can increase coral bleaching [@decarlo2020; @higuchi2015]. These environmental factors influence the resilience of reefs to climate change pressures in combination with related biological factors such as larval connectivity and thermal tolerance.

Connectivity in systems such as the GBR is an important contributor to system resilience, as strong connections are required to repopulate locations after disturbance, and drive adaptation to climate pressures [@angeler2023; @elmhirst2009; @hock2017]. When populations that are strong larval suppliers to external reefs are impacted by disturbances, increased adult coral mortality will cause decreased outgoing larvae in the following years. Weaker larval connections then slow the recovery of the externally connected locations [@figueiredo2022; @hock2017; @munday2009]. Populations with greater strength and number of outgoing than incoming larval connections - herein referred to as larval sources - have lower resilience to pressures than populations with greater incoming than outgoing larval flows - referred to as larval sinks - as larval source populations lack external sources of larvae that are important in contributing to reef recovery and recruitment of juveniles after disturbances to existing populations [@ani2024; @hock2017]. Larval source populations are an important component of overall reef resilience as they can contribute larvae to recovering populations at other locations [@hock2017; @thomas2015]. Connectivity will be altered by future climate change impacts such as population fragmentation, increasing larval mortality, and changes to coastal hydrodynamics [@figueiredo2022; @munday2009], and influences the biology of surrounding reef populations.

Many additional biological factors including coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and acclimation and inherited thermal tolerance[@humanes2022; @lachs2024], influence the outcome of high coral heat stress. It has been shown that corals have the adaptive capacity to develop thermal tolerance levels capable of delaying the onset of widespread bleaching under low emissions scenarios [@bairos-novak2021; @humanes2022].

Scleractinian corals are primary habitat structures on coral reefs. The accumulation and erosion of coral reef carbonate frameworks is controlled by a range of processes such as coral accumulation *Perry*, bioerosion *Tribollet*, physical damage generating coral rubble *Kenyon*, and calcificationa and *CaCO3* dissolution rates *Eyre 2014, 2018*. The balance of these processes and occurrance of net carbonate framework production (referred to as a positive carbonate budget) are influenced by factors such as coral cover and complexity *Perry, Desbiens*, rates of physical environmental processes and ocean acidity *Perry, Eyre*. It has been suggested that total live coral cover on a reef highly influences the maintainence of net positive carbonate budgets. Estimates of live coral cover required to maintain net positive carbonate budget levels range from 10 to 20% *Perry* of a reef's total area, with a level of 17% live coral cover for mid and outershelf reefs on the Great Barrier Reef *Desbiens*.

Incorporating the latest advancements in whole ecosystem larval connectivity modelling and coral thermal response and adaptation research into coral population modelling allows us to investigate how abiotic and biotic factors interact and will influence coral climate change resilience and the maintainence of positive carbonate budgets into the future. Improvements in coral reef climate response modelling, including sampling from a range of possible climate futures, allow for fine-scale timeseries analyses of uncertain climate change and ecological scenarios [@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015].

It is hypothesised that sink reefs are more likely to be resilient under climate change, compared to source reefs, due to greater influx of coral recruits post disturbance. We predict that increasing connectivity levels will correlate with increasing maintainence of positive carbonate budgets. It is hypothesised that reefs with greater depths are more likely to be resilient under future thermal stress, compared to reefs at shallow depths, as they benefit from thermal refugia. We predict that increasing reef depth will correlate with increasing maintainence of positive carbonate budgets. It is hypothesised that reef cover will decline by greater magnitude as the Equilibrium Climate Sensitivity (ECS) of each Global Climate Model (GCM) increases.

@@

@@

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas. It follows the reef definitions as adopted in ReefMod-GBR [@bozec2022] (@fig-region-map). Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon]. Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology [@liu2017; @skirving2020]. These GCMs were selected from available GCMs downscaled by Dixon et al. [-@dixon] based on evaluation by Grose et al. [-@grose2023], which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration [@forster2021]. The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023; @shepherd2018]). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020; IPCC -@climate2023].

@@

## Coral reef ecosystem model - ADRIA x CoralBlox

@@

We track the thermal tolerance of each functional group and size class as a truncated normal distribution that is updated in each timestep following growth and recruitment. This distribution determines that proportion of the population on a reef that is susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value). Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al. [-@bairos-novak2021] and unpublished data from Hughes et al. [-@hughes2018].

Tropical cyclones and wave impacts to reefs are not included in this investigation. We scale coral bleaching impacts by a coefficient determined by the median depth of each reef location, following Baird et al. [@baird2018], reducing the proportion of coral colonies on a reef impacted by high heat stress as depth increases. Median depths for each reef location have been extracted from satellite derived bathymetry data for the Great Barrier Reef (GBRMPA [-@gbr10gb2020]) (see discussion for more details). Due to varying degrees of alignment between bathymetry data and the defined outlines for GBR reefs, reefs with less than 5% coverage of valid bathymetry data have been removed from this investigation. Furthermore, reefs that are not allocated to a GBRMPA management area or bioregion have been removed from the investigation, as well as reefs with relative coral cover values greater than five.

@@

To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries [@batista2014]. These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [@steinmann2020].

Timeseries clustering analyses were performed at the reefal bioregion scale to capture groups of biogeographically similar reefs, as determined by *insert citation for bioregions*. Analyses were also performed at higher spatial scales, GBRMPA management regions and whole GBR scale. Similar results were found at large spatial scales, these can be found at *insert citation for repository*.

Reefs in each cluster are compared to identify characteristics that differ between low, medium and high coral cover clusters. These characteristics include reef median depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation between total coral cover and inversed DHW timeseries, proportion of coral habitable area occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and total strength of larval connections between reefs. To analyse the DHW - tolerance levels of reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep * taxa * reef * scenario matrix. The mean timeseries was calculated over the first four scenarios for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then used to calculate an overall mean DHW – tolerance value for each reef under each GCM.

@@

Overall, our results suggest that when modelled as part of a large and complex ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in support of depth as a form of refugia for shallow reefs. While protecting ‘key node’ reefs in larval connectivity networks is an important consideration for reef restoration efforts [@anthony2020; @hock2017], our results suggest that resilient reefs do not have greater incoming or total connectivity to contribute to their resilience. Instead, we suggest that depth, and likely other environmental thermal refugia mechanisms contribute greater to overall reef resilience and recovery. We suggest further analyses that investigate the benefits of incorporating depth-based resilience into reef decision making.

>>>>

Climate change is impacting species survival, persistence and ecosystem functions globally, including coral reefs. Coral reefs are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events linked to climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral reefs are, and will continue to, decrease the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well as ecosystem services such as food, protection, tourism and Indigenous cultural value [@eddy2021; @stoeckl2021]. We aim to assess the impacts of possible climate futures and identify the characteristics of resilient reefs using the latest generation of larval connectivity research with scenario-based population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. Current modelling project severe declines in coral cover under projections of Degree Heating Weeks [https://www.biorxiv.org/content/10.1101/2025.01.23.634487v1]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

Connectivity is said to play a crucial role in Great Barrier Reef resilience by enabling recolonization after disturbances and facilitating adaptation to climate pressures [@angeler2023; @elmhirst2009; @hock2017]. This connectivity, however, creates both strengths and vulnerabilities within the system. Larval source populations (those exporting more larvae than they import) are essential for system-wide resilience, as they provide larvae that enable recovery at distant locations after disturbances [@hock2017; @thomas2015]. Individually, however, they are more vulnerable to extreme disturbances as they rely more heavily on local recruitment to support recovery. In contrast, sink populations can draw on external larval inputs from multiple and diverse sources [@ani2024; @hock2017]. When disturbances impact major larval source populations, the cascading effects extend throughout the network. Increased adult mortality reduces larval export in subsequent years, slowing recovery at the connected sink locations that depend on this external supply [@figueiredo2022; @hock2017; @munday2009]. Future climate change will further complicate these dynamics through population fragmentation, increased larval mortality, and altered hydrodynamics, potentially disrupting the connectivity patterns that currently maintain reef resilience.

Beyond connectivity, coral resilience to climate change is determined by the complex interplay of environmental and biological factors. Environmental abiotic factors influence how reefs will experience climate change pressures, including the spatio-temporal patterns of thermal stress itself, and indirect factors that influence the stress experienced by corals. Environmental factors such as tropical cyclones and wind driven mixing [@lin2023; @mcclanahan2005]; currents [@grimaldi2022] and depth [@álvarez-noriega2018; @baird2018; @smith2014] can decrease the magnitude of thermal stress corals experience. Conversely, high light intensity and nutrient levels in the benthic zone can increase coral bleaching [@decarlo2020; @higuchi2015].

Individual coral resilience also depends on multiple biological factors beyond connectivity. Coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and acclimation and inherited thermal tolerance [@humanes2022; @lachs2024] all influence the outcome of high coral heat stress events. Importantly, corals have demonstrated adaptive capacity to develop thermal tolerance levels capable of delaying the onset of widespread bleaching under low emissions scenarios [@bairos-novak2021; @humanes2022]. This adaptive potential represents a crucial mechanism for reef persistence under moderate climate change, working in concert with the environmental factors that modulate thermal stress exposure.

While these environmental and biological factors determine whether individual corals survive thermal stress events, the long-term viability of coral reefs depends not only on coral survival but also on the maintenance of reef structure through positive carbonate budgets. Scleractinian corals are primary habitat structures on coral reefs, and the accumulation and erosion of coral reef carbonate frameworks is controlled by a range of processes including coral accumulation [**Perry**], bioerosion [**Tribollet**], physical damage generating coral rubble [**Kenyon**], and calcification and CaCO₃ dissolution rates [**Eyre 2014, 2018**]. The balance of these processes and occurrence of net carbonate framework production (referred to as a positive carbonate budget) are influenced by factors such as coral cover and complexity [**Perry, Desbiens**], rates of physical environmental processes, and ocean acidity [**Perry, Eyre**]. Total live coral cover on a reef highly influences the maintenance of net positive carbonate budgets, with estimates ranging from 10 to 20% [**Perry**] of a reef's total area required to maintain positive budgets. For mid and outer shelf reefs on the Great Barrier Reef, this threshold is approximately 17% live coral cover [**Desbiens**].

Incorporating the latest advancements in whole ecosystem larval connectivity modeling and coral thermal response and adaptation research into coral population modeling allows an investigation into how abiotic and biotic factors interact to influence coral climate change resilience and the maintenance of positive carbonate budgets. Improvements in coral reef climate response modeling, including sampling from a range of possible climate futures, enable timeseries analyses of uncertain climate change and ecological scenarios [@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015]. While it has previously been suggested that highly connected source reefs that are resilient to bleaching and CoTS outbreaks are likely to play a critical role in supporting future reef recovery across the entire GBR [@hock2017], integrated modelling under current projections of Degree Heating Weeks indicate that depth, rather than connectivity, plays the critical role in determining reef resilience. It is further demonstrated that depth ameliorates reef cover declines and supports maintenance of a postive carbonate budget as the Equilibrium Climate Sensitivity (ECS) of each Global Climate Model (GCM) increases.

@@

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas. It follows the reef definitions as provided by the Great Barrier Reef Marine Park Authority [https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::great-barrier-reef-features-/explore?location=-19.221541%2C145.776418%2C4.94]. Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6; **citation**) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon]. Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology [@liu2017; @skirving2020]. These GCMs were selected from available GCMs downscaled by Dixon et al. [-@dixon] based on evaluation by Grose et al. [-@grose2023], which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration [@forster2021]. The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023; @shepherd2018]). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020; IPCC -@climate2023].

@@

@@

## Coral reef ecosystem model - ADRIA-CoralBlox

@@

We track the thermal tolerance of each functional group and size class as a truncated normal distribution that is updated in each timestep following growth and recruitment. This distribution determines that proportion of the population on a reef that is susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value). Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al. [-@bairos-novak2021].

Tropical cyclones and impact of wave activity to reefs are not included in this investigation. We scale coral bleaching impacts by a coefficient determined by the median depth of each reef location, following Baird et al. [@baird2018], reducing the proportion of coral colonies on a reef impacted by high heat stress as depth increases. Median depths for each reef location have been extracted from satellite derived bathymetry data for the Great Barrier Reef (GBRMPA [-@gbr10gb2020]) (see discussion for more details). Due to varying degrees of alignment between bathymetry data and the defined outlines for GBR reefs, reefs with less than 5% coverage of valid bathymetry data have been removed from this investigation. Furthermore, reefs that were outside the GBRMPA management areas or not associated with a bioregion were removed from the investigation, as well as reefs with relative coral cover values greater than five [**five what???**].

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we clustered reefs by their temporal behaviour as indicated by their coral cover over time. The Complexity-Invariant Distance (CID) between each timeseries was calculated to identify timeseries that behave similarly [@batista2014]. The CID corrects the Euclidean distance between pairs of timeseries (which accounts for magnitude differences) with the complexity of each timeseries (obtained through a measure of their variability) to ensure reefs are clustered based on their underlying temporal dynamics rather than simply their average coral cover. The process creates a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020].

Timeseries clustering analyses were performed at the reefal bioregion scale (defined in [https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::reefal-marine-bioregions/about]).
Additional analyses were also performed at larger spatial scales, such as the GBRMPA management regions and whole-of-GBR scale. These were found to lead to identical conclusions and so are not shown here.
Results of these analyses can be found at [**insert citation for repository**] for those interested.

Reefs in each timeseries cluster are compared to identify characteristics that differ between low, medium and high coral cover clusters. These characteristics include reef median depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation between total coral cover and inversed DHW timeseries, proportion of coral habitable area occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and total strength of larval connections between reefs. To analyse the DHW - tolerance levels of reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep * taxa * reef * scenario matrix. The mean timeseries was calculated over the first four scenarios for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then used to calculate an overall mean DHW – tolerance value for each reef under each GCM.

@@

Overall, our results suggest that when modelled as part of a large and complex ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in support of depth as a form of refugia for shallow reefs. While protecting ‘key node’ reefs in larval connectivity networks is an important consideration for reef restoration efforts [@anthony2020; @hock2017], our results suggest that resilient reefs do not have greater incoming or total connectivity to contribute to their resilience under modelled projected conditions. Instead, we suggest that depth, and likely other environmental thermal refugia mechanisms contribute greater to overall reef resilience and recovery. We suggest further analyses that investigate the benefits of incorporating depth-based resilience into reef decision making.
